---
title: "Homework 4 Math 6359"
author: "Justin Mundell"
date: "2024-04-10"
output:
  word_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

# Data Exploration

We will start reading the data. We will do some column names adjustments
to have a better readability.

We will explore the dimensions of each data frame we will be working
with. Also, we will look into Type 1 unique values as we are interested
in comparing the performance of the Pokemon based on what type they
are.

Finally, we will make sure there are no missing values.

```{r}
pokemon = read.csv("C:/Users/Handt/OneDrive/Documents/Math 6359/Project/pokemon.csv")
names(pokemon)[names(pokemon) == 'X.'] <- 'Number'
```

```{r}
combats = read.csv("C:/Users/Handt/OneDrive/Documents/Math 6359/Project/combats.csv")
```

```{r}
head(pokemon)
dim(pokemon) #there are 800 Pokemon
```

```{r}
head(combats)
dim(combats) #there are approximately 50,000 battles between Pokemon
```

```{r}
unique(pokemon$Type.1) ; length(unique(pokemon$Type.1)) #number of types
```

```{r}
colSums(is.na(pokemon)) #there are no missing values
```

# Data Preparation

#### Win Percentage Calculation We will first calculate the win percentage of 
#### each Pokemon and create a column showing this.

```{r}
first_count = table(combats$First_pokemon)
second_count = table(combats$Second_pokemon)

total_battles = first_count + second_count #total battles per Pokemon

total_wins = table(combats$Winner) # total wins per Pokemon
```

```{r}
length(as.vector(total_wins))
length(as.vector(total_battles))

total_battles = as.data.frame(total_battles)
total_wins = as.data.frame(total_wins)
```

The number of dimensions are different in total wins and total battles.
This indicates that there is one Pokemon that did not win any of its
battles. We must find this Pokemon and add it to total wins to correct
the dimensions.

```{r}
setdiff(total_battles$Var1, total_wins$Var1)
print(pokemon[231,])
```

The Pokemon that did not win any battles is Shuckle.

```{r}
shuckle <- data.frame(Var1 = 231, Freq = 0)
total_wins <- rbind(shuckle, total_wins)

colnames(total_battles) = c("Number", "Battles")
colnames(total_wins) = c("Number", "Wins")
```

```{r}
#Calculating win percentage
df = merge(total_battles,total_wins, by = "Number")
df$`Win Percentage` = df$Wins/df$Battles
```

```{r}
pokemon <- merge(pokemon, df, by = "Number")
```

#### Top Pokemon of each Type (18 best based on Type1) This shows the Pokemon 
#### that has the highest winning percentage for each Pokemon type.
```{r}
library(dplyr)
best_pokemon_type1 <- pokemon %>%
  arrange(desc(`Win Percentage`)) %>%
  group_by(Type.1) %>%
  slice(1)
best_pokemon_type1 
```

# Analysis

## Pokemon comparisons on having Type 1 only (18 unique types).

Is there a statistically significant difference in the mean win
percentage among Pokemon that only have one type?

There are 18 total types.

```{r}
two_types = pokemon[Reduce(`&`, 
                           lapply(pokemon, function(x) !(is.na(x)|x==""))),]

one_type = pokemon[!Reduce(`&`, 
                           lapply(pokemon, function(x) !(is.na(x)|x==""))),]

two_types = two_types[, c("Number", "Name", "Type.1", "Type.2", 
                         "Win Percentage")]

one_type = one_type[, c("Number", "Name", "Type.1",  "Win Percentage")]
```

#### Check Assumptions

```{r}
library(car)
library(biotools)

# Normality
for(type in unique(one_type$Type.1)) {
  subset <- one_type[one_type$Type.1 == type,]
  
  # Check if the sample size is within the valid range for Shapiro-Wilk test
  if(nrow(subset) >= 3 & nrow(subset) <= 5000) {
    shapiro_test <- shapiro.test(subset$`Win Percentage`)
    print(paste("Type:", type, "p-value:", shapiro_test$p.value))
  } else {
    print(paste("Type:", type, "invalid sample size for Shapiro-Wilk test"))
  }
}

# Homogeneity of variances
levene_test <- leveneTest(`Win Percentage` ~ Type.1, data = one_type)
print(levene_test)

```

From our results, it appears that the win percentages for the types Bug,
Normal, Electric, Psychic, and Grass have p-values less than 0.05,
suggesting that they do not follow a normal distribution.

The type Flying has an invalid sample size for the test since there are a total
of 2 flying Pokemon. There needs to be at least 3 per group for the test to
work.

```{r}
sum(one_type$Type.1 == "Flying") # number of Pokemon that fly
```
From our results, the p-value is 0.001667, which is less than 0.05. This
suggests that we reject the null hypothesis and conclude that the
variances are not equal across the different Pokemon types.

#### Testing if there is a significant statistical difference in win percentage

Our first idea was to use ANOVA but because our data does no meet the
assumptions of normality and homogeneity of variances we will look into
a non-parametric version of the test, such as Kruskal-Wallis.

#### ANOVA

```{r}
#ANOVA function for pokemon with one type
attach(one_type)

one_type$Type.1 <- as.factor(one_type$Type.1)
one.lm = lm(`Win Percentage`~ Type.1, data = one_type)
anova(one.lm)
```

There's evidence that at least one of the mean win percentages among the
types of Pokemon are not equal.

```{r}
#Pairs t-test
pairwise.t.test(`Win Percentage`, Type.1, "bon")
```

According to the pairwise T test, there's a statistical significance
difference between the bug type versus dark type, bug versus electric
type, bug versus fire type, and fairy versus electric type. This is
telling us that the bug type is the one that's mostly significantly
different from all of the other ones.

This indicates that the bug type Pokemon loses a lot of battles.

```{r}
library(tidyverse)
qplot(data = one_type, x = Type.1, y = `Win Percentage`, geom = "boxplot")
```

This box plot provides further reasoning to why the bug type Pokemon has
a very low win percentage out of all Pokemon types. It's important to note that 
the mean win percentage of the flying type Pokemon is extremely high since there
are only 2 observations of that type. Thus, it's an outlier.

#### Kruskal-Wallis

```{r}
kruskal_test <- kruskal.test(`Win Percentage` ~ Type.1, data = one_type)
print(kruskal_test)
```

The p-value we obtained is less than 0.05, which suggests that we can
reject the null hypothesis and conclude that there is a statistically
significant difference in the Win Percentage across the different
Pokemon types.

However, the Kruskal-Wallis test does not tell us which specific groups
(in this case, Pokemon types) were significantly different from each
other. To find out, we can perform the Dunn test.

```{r}
library(dunn.test)
dunn_result <- dunn.test(one_type$`Win Percentage`, one_type$Type.1, 
                         method = "bonferroni")
```

The results from the Dunns test indicate that there are statistically
significant differences in win percentages between Bug type and Dark,
Electric, and Fire type.

```{r}
# Calculate median win percentage for Bug type
bug_median <- median(one_type$`Win Percentage`[one_type$Type.1 == "Bug"])

# Calculate median win percentage for Dark type
dark_median <- median(one_type$`Win Percentage`[one_type$Type.1 == "Dark"])

# Calculate median win percentage for Electric type
electric_median <- median(one_type$`Win Percentage`[one_type$Type.1 == "Electric"])

# Calculate median win percentage for Fire type
fire_median <- median(one_type$`Win Percentage`[one_type$Type.1 == "Fire"])

# Print the results
print(paste("Median win percentage for Bug type:", bug_median))
print(paste("Median win percentage for Dark type:", dark_median))
print(paste("Median win percentage for Electric type:", electric_median))
print(paste("Median win percentage for Fire type:", fire_median))

```

Based on the median win percentages we have calculated, it appears that
Bug types have a lower median win percentage (0.146) compared to Dark
(0.665), Electric (0.604), and Fire (0.536) types. This suggests that,
on average, Bug type Pokemon tend to win less often when battling
against these types.

#### Analysis on our findings

Initially, we planned to use an ANOVA test to determine if there was a
statistically significant difference in the win percentages across the
different Pokemon types. However, upon checking the assumptions for the
ANOVA test, we found that they were not met. Specifically, the
assumption of normality was violated for several Pokemon types, and the
assumption of homogeneity of variances was also not met.

As a result, we turned to the non-parametric Kruskal-Wallis test, which
does not require these assumptions. The results from the Kruskal-Wallis
test were consistent with what we would have expected from the ANOVA
test, had its assumptions been met. We found that there is a
statistically significant difference in the win percentages across the
different Pokemon types.

Further analysis using Dunn’s post-hoc test revealed that the Bug type
Pokemon showed statistically significant differences in win percentages
when compared to other types. This suggests that the performance of Bug
type Pokemon in battles is significantly different from that of several
other types.

It’s important to note that while the assumptions for the ANOVA test
were not met, leading us to use the Kruskal-Wallis test instead, both
tests led us to the same conclusion about the overall differences in win
percentages across Pokemon types. This demonstrates the robustness of
our findings.

## Pokemon comparisons on having 1 Type versus 2 Types

```{r}
library(ICSNP)

# Convert vectors to one-column matrices
one_type_matrix <- matrix(one_type$`Win Percentage`, ncol = 1)
two_types_matrix <- matrix(two_types$`Win Percentage`, ncol = 1)

# Perform HotellingsT2 test
HotellingsT2(one_type_matrix, two_types_matrix)

```
```{r}
# Perform Mann-Whitney U test
mw_test = wilcox.test(one_type$`Win Percentage`, two_types$`Win Percentage`)
mw_test
```

The results from both the Hotellings T2 test and the Mann-Whitney U test 
indicate that there is a statistically significant difference in the Win 
Percentage between Pokemon with one type and those with two types.

The test statistic for Hotellings T2 is 14.005, and the p-value is 0.0001956. 
Since the p-value is less than 0.05, we can reject the null hypothesis and 
conclude there is a statistically significant difference in the Win Percentage 
between Pokemon with one type and those with two types.

The test statistic for Mann-Whitney is 64853, and the p-value is 0.0001771. 
Again, since the p-value is less than 0.05, we can reject the null hypothesis 
and conclude that there is a statistically significant difference in the Win 
Percentage between Pokemon with one type and those with two types.

In conclusion, both tests suggest that the number of types a Pokemon has 
(one type vs. two types) significantly affects its win percentage. However, 
these tests do not tell us which group (one type or two types) tends to have a
higher win percentage. For that, we could look at the median or mean win 
percentages of the two groups.

*Look at median or mean win percentages of the two groups*

#### This is for pokemon of just one type
```{r}
#The medians of Pokemon with just one type (by type)
unique_groups <- unique(one_type$Type.1)

for (group_name in unique_groups) {
  group_median <- median(one_type$`Win Percentage`[one_type$Type.1 == group_name], na.rm = TRUE)
  cat("Type:", group_name, "- Median Value:", group_median, "\n")
}

```
```{r}
medians <- numeric(length(unique_groups))

# calculating the medians of each single type
for (i in seq_along(unique_groups)) {
  group_name <- unique_groups[i]
  group_median <- median(one_type$`Win Percentage`
                         [one_type$Type.1 == group_name], na.rm = TRUE)
  medians[i] <- group_median
}

# Plot histogram
hist(medians, main = "Histogram of Median Win Percentages by Type", 
     xlab = "Median Win Percentage", col = "red")
```

After calculating the median win percentage for Pokemon of one type and viewing
the histogram, it's difficult to identify a central tendency. The distribution 
is skewed to the right, showing greater variability. This means that there are 
possible outliers. From the output of the medians, the flying type has a median
win percentage of 90.74%. Earlier, we found that there are only two Pokemon in 
the data set that are just the flying type. This explains why the distribution
is extremely skewed.

Furthermore, 9 out of 18 of the types fall within the 30-40% median win 
percentage range. Ice, steel, rock, dragon, grass, fighting, fairy, ground, and
poison types fall within this range. Excluding the flying type, dark, psychic, 
electric, normal, and fire all have a median win percentage above 50%. This goes
to show which Pokemon types win more often than others.


```{r}
one_type_counts <- table(one_type$Type.1)
one_type_counts
```
For further investigation, the Pokemon types that make up most of the population 
are normal, psychic, fire, water, electric, and fighting. Besides the dark type,
these types all have a median win percentage above 50%. Having a higher sample
size for each Pokemon type could explain the robustness of the median win 
percentage, but it may not mean that the win percentage would be higher. Pokemon
types vary in their strengths and weaknesses, which influences their likelihood
of winning in battles. Some types may naturally  have higher win percentages due
to their effectiveness against other types, regardless of sample size.

```{r}
# removing Flying-type Pokémon
one_type_no_flying <- one_type[one_type$Type.1 != "Flying", ]

overall_one_median <- median(one_type_no_flying$`Win Percentage`, 
                             na.rm = TRUE)
print(overall_one_median)
```

```{r}
unique_types <- unique(one_type_no_flying$Type.1)
type_medians <- numeric(length(unique_types))

for (i in seq_along(unique_types)) {
  type <- unique_types[i]
  type_medians[i] <- median(one_type_no_flying$`Win Percentage`
                            [one_type_no_flying$Type.1 == type], na.rm = TRUE)
}

# plotting the distribution of median win percentages
hist(type_medians,
     main = "Median Win Percentages (Excluding Flying Type)",
     xlab = "Median Win Percentage",
     col = "skyblue",
     border = "black")
```

When excluding the two flying type observations and recreating the histogram,
the distribution appears to be less skewed. The distribution is still not 
normal, but for the purpose of comparison, we decided to find the central 
tendency. The overall median of the median win percentages for each type is 
43.38%. 



#### This is for pokemon of two types




```{r}
result <- aggregate(two_types$`Win Percentage`~ two_types$Type.1 + 
                      two_types$Type.2, data = two_types, FUN = median, 
                    na.rm = TRUE)
head(result) #first 10 
```
```{r}
dim(result)
```

```{r}
hist(result$`two_types$\`Win Percentage\``, breaks = 10, col = "lightblue",
     main = "Distribution of Median Win Percentage for Two Types",
     xlab = "Median", ylab = "Frequency")
```

There are 136 total unique combinations for Pokemon that have two
types. This means that the number of observations per combination may vary. 
However, there seems to be a normal distribution of the median win percentage 
for Pokemon of two types. 

Since the distribution of the medians are normal, it's appropriate to calculate
the central tendency of the medians.

```{r}
median_win_percentages <- result$`two_types$\`Win Percentage\``

overall_median <- median(median_win_percentages, na.rm = TRUE)

# Print the result
print(overall_median)
```

The overall median win percentage of Pokemon with two types is 55.36%. This 
shows that the average performance of Pokemon across various types. Thus, since 
the median is over 50%, it suggests that on average, Pokemon with type 
combinations are relatively balanced and perform well. This is much higher than
the overall median win percentage for Pokemon that have just one type, which was
43.38%. 

From the Hotellings T2 test and the Mann-Whitney U test, we found that there's
a statistically significant difference between the Pokemon of one type and 
Pokemon of two types. Finding the overall median win percentage for the both
of them has shown that there is a difference in their overall performances,
which supports the findings from these tests.

## Pokemon comparisons on different generations

#### Check Assumptions

```{r}
pokemon$Generation <- as.factor(pokemon$Generation)

# Normality
for(generation in unique(pokemon$Generation)) {
  subset <- pokemon[pokemon$Generation == generation,]
  
  # Check if the sample size is within the valid range for Shapiro-Wilk test
  if(nrow(subset) >= 3 & nrow(subset) <= 5000) {
    shapiro_test <- shapiro.test(subset$`Win Percentage`)
    print(paste("Generation:", generation, "p-value:", shapiro_test$p.value))
  } else {
    print(paste("Generation:", generation, "has an invalid sample size for Shapiro-Wilk test"))
  }
}

# Homogeneity of variances
levene_test <- leveneTest(`Win Percentage` ~ Generation, data = pokemon)
print(levene_test)

```

For some background information the generation is which season the pokemon came 
out. (this should go in introduction***)

From our results, it appears that the win percentages for all Pokemon 
generations except Generation 6 have p-values less than 0.05, suggesting 
that they do not follow a normal distribution. For Generation 6, the p-value is 
slightly greater than 0.05. Thus, we reject the null hypothesis and conclude 
that the generations do not follow a normal distribution.

From our results, the p-value is 0.7527, which is greater than 0.05. We fail to
reject the null hypothesis and conclude that the variances are equal across the 
different Pokemon generations.

#### Testing if there is a significant statistical difference in win percentage

As we have found from our previous tests, the win percentages for all Pokemon 
generations except Generation 6 do not follow a normal distribution, which 
violates one of the key assumptions of ANOVA. However, the variances across the
different Pokemon generations are equal, which meets the homogeneity of 
variances assumption of ANOVA.

Given these findings, the Kruskal-Wallis test, which does not require the 
assumption of normality, would be a more appropriate choice in this case.

#### Kruskal-Wallis

```{r}
kruskal_test <- kruskal.test(`Win Percentage` ~ Generation, data = pokemon)
print(kruskal_test)
```

The p-value is 0.08518, which is greater than the commonly used significance 
level of 0.05. This means that we fail to reject the null hypothesis. 
Our results do not provide strong evidence to conclude that there are 
statistically significant differences in the Win Percentage across the different 
Pokemon generations. 

```{r}
ggplot(pokemon, aes(x = Generation, y = `Win Percentage`)) +
  geom_boxplot(fill = "yellow", color = "black") +
  labs(title = "Win Percentage by Generation",
       x = "Generation",
       y = "Win Percentage")
```


```{r}
generation_medians <- aggregate(`Win Percentage` ~ Generation, data = pokemon, 
                                FUN = median)
print(generation_medians) # median win percentage for each distribution
```

This can also been seen in the box plot that shows the median win percentage for
each generation. The win percentages for each generation are 0.52, 0.43, 0.43,
0.59, 0.48, and 0.46 respective to their generations. Although the fourth 
generation has the highest win percentage, there isn't a large enough difference 
in these values to suggest there's a significant difference. 

Lastly, these results suggests that the generation a Pokemon is from doesn't 
necessarily impact their likelihood of winning or losing a battle.


## Pokemon comparisons on being legendary

#### Check Assumptions

```{r}
pokemon$Legendary <- as.factor(pokemon$Legendary)

# Normality
legendary_pokemons <- pokemon[pokemon$Legendary == "True",]
non_legendary_pokemons <- pokemon[pokemon$Legendary == "False",]

shapiro_test_leg <- shapiro.test(legendary_pokemons$`Win Percentage`)
shapiro_test_non_leg <- shapiro.test(non_legendary_pokemons$`Win Percentage`)

print(paste("Legendary Pokemons p-value:", shapiro_test_leg$p.value))
print(paste("Non-Legendary Pokemons p-value:", shapiro_test_non_leg$p.value))

# Homogeneity of variances
levene_test <- leveneTest(`Win Percentage` ~ Legendary, data = pokemon)
print(levene_test)

```

From our Shapiro Test results, it appears the win percentages for both Legendary 
and Non-Legendary Pokemon have p-values less than 0.05, suggesting that they do 
not follow a normal distribution.

After testing the homogeneity of the variances, we found that the p-value is 
significantly less than 0.05. We reject the null hypothesis. This suggests the 
variances are not equal between Legendary and Non-Legendary Pokemon.

Given these findings, the assumptions for an ANOVA test are not met. 
Therefore, a non-parametric test like the Mann-Whitney U test would be more 
appropriate for comparing the Win Percentage between Legendary and 
Non-Legendary Pokemon.

#### Mann-Whitney

```{r}
# Perform Mann-Whitney U test
wilcox.test(legendary_pokemons$`Win Percentage`, non_legendary_pokemons$`Win Percentage`)

```

The p-value is less than 2.2e-16, which is essentially zero and much smaller 
than the significance level of 0.05. We reject the null hypothesis and conclude
there is very strong evidence of a statistically significant difference in the 
Win Percentage between Legendary and Non-Legendary Pokemon.

If we look at the Legendary and Non-Legendary Pokemon win percentage:

```{r}
print(paste("Legendary Pokemons Win Percentage median:", 
            median(legendary_pokemons$`Win Percentage`)))

print(paste("Non-Legendary Pokemons Win Percentage median:", 
            median(non_legendary_pokemons$`Win Percentage`)))
```

We can observe that Legendary Pokemon have a median win percentage of 
approximately 0.821, while Non-Legendary Pokemon have a median win percentage of
about 0.456. This suggests that Legendary Pokemon win battles almost twice as 
often as Non-Legendary Pokemon. This could be due to the typically superior 
stats and abilities of Legendary Pokemon.

```{r}
ggplot(pokemon, aes(x = Legendary, y = `Win Percentage`, fill = Legendary)) +
  geom_boxplot() +
  labs(title = "Median Win % for Legendary and Non-Legendary Pokémon",
       x = "Legendary Status",
       y = "Win Percentage",
       fill = "Legendary") +
  theme_minimal() 
```
